\bf Appendix: Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-05-14 appendix.qmd
Started updating the appendix

scurve_scaled_obj <- gen_scaled_data(
  data = umap_scurve)

umap_scurve_scaled <- scurve_scaled_obj$scaled_nldr
lim1 <- scurve_scaled_obj$lim1
lim2 <- scurve_scaled_obj$lim2
r2 <- diff(lim2)/diff(lim1)

@@

## hexagon binning to have regular hexagons
hb_obj_scurve <- hex_binning(
  data = umap_scurve_scaled, 
  bin1 = num_bins_x_scurve, 
  r2 = r2)

@@

```{r}
# Code to draw illustration for notation
start_pt <- all_centroids_df |> 
  filter(hexID == 1)
d_rect <- tibble(x1min = 0, 
                 x1max = 1,
                 x2min = 0,
                 x2max = r2)

@@

$$
\frac{2(r_2 + qr_2)}{\sqrt{3}(b_2 - 1)} = \frac{(1 + q)}{(b_1 - 1)}.
$$ After rearranging this,

@@

## hexagon binning to have regular hexagons
hb_obj_scurve <- hex_binning(
  data = umap_scurve_scaled, 
  bin1 = num_bins_x_scurve, 
  r2 = r2)

@@

```{r}
#| label: fig-assign-data
#| fig-cap: "Binning the data. Points are assigned to the nearest centroid. If a point is equidistant from multiple centroids, assigned to the lowest centroid."
#| fig-width: 9
#| fig-height: 6
assign_data_vis1 + assign_data_vis2 +
  plot_layout(ncol = 2)
```

@@

### Single-cell gene expression

#### Comparison with results of scDEED recommendations

@@

error_pbmc <- error_pbmc |>
  mutate(a1 = round(a1, 2)) |>
  filter(bin1 >= 5) |>
  group_by(method, a1) |>
  filter(MSE == min(MSE)) |>
  ungroup()
```

@@

```{r}
#| label: fig-pbmc-mse-umap
#| fig-pos: H
#| out-width: 100%
#| fig-cap: "Assessing which of the two layouts with UMAP and tSNE with different hyper-parameter setting (n_neighbors: $30$, min_dist: $0.3$ (red); n_neighbors: $30$ (blue)) on the PBMC data is the better representation using MSE for varying binwidth ($a_1$). Colour used for the lines and points in the top plot and in the scatterplots represents UMAP and tSNE layouts (a, b). Of the two, layout b is optimal across all binwidths making it the best choice."

@@

algo_obj_pbmc <- gen_nldr_vis_algo_obj(
  high_d_data = training_data_pbmc, 
  nldr_data = umap_pbmc, 
  num_x_bins = num_bins_x_pbmc)

umap_pbmc_scaled <- algo_obj_pbmc$nldr_scaled
distance_pbmc <- algo_obj_pbmc$distance_df
tr_from_to_df_pbmc <- algo_obj_pbmc$tr_from_to_df
benchmark_pbmc <- algo_obj_pbmc$benchmark
df_bin_centroids_pbmc <- algo_obj_pbmc$df_bin_centroids
df_bin_pbmc <- algo_obj_pbmc$df_bin

distance_df_small_edges_pbmc <- distance_pbmc |>
  filter(distance < benchmark_pbmc)#benchmark_pbmc

@@

tr_from_to_df_pbmc <- right_join(
  tr_from_to_df_pbmc, distance_df_small_edges_pbmc, 
  by = c("from", "to"))

@@

langevitour::langevitour(df_model_data_pbmc_n[1:(length(df_model_data_pbmc_n)-1)],
                         lineFrom = distance_df_small_edges_pbmc$from,
                         lineTo = distance_df_small_edges_pbmc$to,
                         group = factor(df_model_data_pbmc_n$type,
                                        c("cluster1", "cluster2", "cluster3", "model")),
                         levelColors = c("#8dd3c7", "#fdb462", "#bebada", "#000000"))
```

@@

proj_obj1 <- get_projection(projection = projection, 
               proj_scale = 1.23, 
               scaled_data = scaled_pbmc_data, 
               scaled_data_model = scaled_pbmc_data_model, 
               distance_df_small_edges = distance_df_small_edges_pbmc, 
               axis_param = list(limits = 1,
                                 axis_scaled = 1, 
                                 axis_pos_x = -0.48, 
                                 axis_pos_y = -0.48, 
                                 threshold = 0.082))

@@

proj_obj2 <- get_projection(projection = projection, 
               proj_scale = 1.25, 
               scaled_data = scaled_pbmc_data, 
               scaled_data_model = scaled_pbmc_data_model, 
               distance_df_small_edges = distance_df_small_edges_pbmc, 
               axis_param = list(limits = 1.35, 
                                 axis_scaled = 1, 
                                 axis_pos_x = -0.75, 
                                 axis_pos_y = -0.75, 
                                 threshold = 0.12))

@@

algo_obj_pbmc <- gen_nldr_vis_algo_obj(
  high_d_data = training_data_pbmc, 
  nldr_data = tsne_pbmc, 
  num_x_bins = num_bins_x_pbmc)

tsne_pbmc_scaled <- algo_obj_pbmc$nldr_scaled
distance_pbmc <- algo_obj_pbmc$distance_df
tr_from_to_df_pbmc <- algo_obj_pbmc$tr_from_to_df
benchmark_pbmc <- algo_obj_pbmc$benchmark
df_bin_centroids_pbmc <- algo_obj_pbmc$df_bin_centroids
df_bin_pbmc <- algo_obj_pbmc$df_bin

distance_df_small_edges_pbmc <- distance_pbmc |>
  filter(distance < benchmark_pbmc)#benchmark_pbmc

@@

pbmc_proj_tsne_model2_best <- plot_proj(
  proj_obj = proj_obj2, 
  point_param = c(1.5, 0.2), # size, alpha, color
  plot_limits = c(-1, 0.75), 
  title = "b3", 
  cex = 2, 
  axis_text_size = 4,
  is_color = TRUE) + scale_color_manual(values = c("#8dd3c7", "#fdb462", "#bebada"))

>>>>

scurve_scaled_obj <- gen_scaled_data(
  nldr_data = umap_scurve)

@@

## hexagon binning to have regular hexagons
hb_obj_scurve <- hex_binning(
  nldr_obj = scurve_scaled_obj, 
  bin1 = 8, 
  q = 0.1)

@@

```{r}
# Code to draw illustration for notation
start_pt <- all_centroids_df |> 
  filter(hexID == 1)
d_rect <- tibble(x1min = 0, 
                 x1max = 1,
                 x2min = 0,
                 x2max = diff(scurve_scaled_obj$nldr_obj$lim2)/diff(scurve_scaled_obj$nldr_obj$lim1))

@@

$$
\frac{2(r_2 + qr_2)}{\sqrt{3}(b_2 - 1)} = \frac{(1 + q)}{(b_1 - 1)}.
$$

After rearranging this,

@@

## hexagon binning to have regular hexagons
hb_obj_scurve <- hex_binning(
  nldr_obj = scurve_scaled_obj, 
  bin1 = num_bins_x_scurve, 
  q = 0.1)

@@

```{r}
#| label: fig-assign-data
#| fig-cap: "Binning the data. Points are assigned to the nearest centroid. If a point is equidistant from multiple centroids, assigned to the lowest centroid."
#| fig-width: 9
#| fig-height: 6

assign_data_vis1 + assign_data_vis2 +
  plot_layout(ncol = 2)
```

@@

## Single-cell gene expression: comparison with results of scDEED recommendations

<!-- ### Comparison with results of scDEED recommendations -->

@@

error_pbmc <- error_pbmc |>
  mutate(a1 = round(a1, 2)) |>
  filter(bin1 >= 5) |>
  group_by(method, a1) |>
  filter(RMSE == min(RMSE)) |>
  ungroup()
```

@@

```{r}
#| label: fig-pbmc-mse-umap
#| fig-pos: H
#| out-width: 100%
#| fig-cap: "Assessing which of the two layouts with UMAP and tSNE with different hyper-parameter setting (n_neighbors: $30$, min_dist: $0.3$ (red); n_neighbors: $30$ (blue)) on the PBMC data is the better representation using RMSE for varying binwidth ($a_1$). Colour used for the lines and points in the top plot and in the scatterplots represents UMAP and tSNE layouts (a, b). Of the two, layout b is optimal across all binwidths making it the best choice."

@@

algo_obj_pbmc <- fit_highd_model(
  highd_data = training_data_pbmc, 
  nldr_data = umap_pbmc, 
  bin1 = num_bins_x_pbmc, 
  q = 0.1, 
  benchmark_highdens = 5)

umap_pbmc_scaled <- algo_obj_pbmc$nldr_obj$scaled_nldr
tr_from_to_df_pbmc <- algo_obj_pbmc$trimesh_data
df_bin_centroids_pbmc <- algo_obj_pbmc$model_2d
df_bin_pbmc <- algo_obj_pbmc$model_highd

@@

@@

langevitour::langevitour(df_model_data_pbmc_n[1:(length(df_model_data_pbmc_n)-1)],
                         lineFrom = tr_from_to_df_pbmc$from,
                         lineTo = tr_from_to_df_pbmc$to,
                         group = factor(df_model_data_pbmc_n$type,
                                        c("cluster1", "cluster2", "cluster3", "model")),
                         levelColors = c("#8dd3c7", "#fdb462", "#bebada", "#000000"))
```

@@

proj_obj1 <- get_projection(projection = projection, 
               proj_scale = 1.23, 
               highd_data = scaled_pbmc_data, 
               model_highd = scaled_pbmc_data_model, 
               trimesh_data = tr_from_to_df_pbmc, 
               axis_param = list(limits = 1,
                                 axis_scaled = 1, 
                                 axis_pos_x = -0.48, 
                                 axis_pos_y = -0.48, 
                                 threshold = 0.082))

@@

proj_obj2 <- get_projection(projection = projection, 
               proj_scale = 1.25, 
               highd_data = scaled_pbmc_data, 
               model_highd = scaled_pbmc_data_model, 
               trimesh_data = tr_from_to_df_pbmc, 
               axis_param = list(limits = 1.35, 
                                 axis_scaled = 1, 
                                 axis_pos_x = -0.75, 
                                 axis_pos_y = -0.75, 
                                 threshold = 0.12))

@@

algo_obj_pbmc <- fit_highd_model(
  highd_data = training_data_pbmc, 
  nldr_data = tsne_pbmc, 
  bin1 = num_bins_x_pbmc, 
  q = 0.1, 
  benchmark_highdens = 1)

tsne_pbmc_scaled <- algo_obj_pbmc$nldr_obj$scaled_nldr
tr_from_to_df_pbmc <- algo_obj_pbmc$trimesh_data
df_bin_centroids_pbmc <- algo_obj_pbmc$model_2d
df_bin_pbmc <- algo_obj_pbmc$model_highd

@@

pbmc_proj_tsne_model2_best <- plot_proj(
  proj_obj = proj_obj2, 
  point_param = c(1.5, 0.2), # size, alpha, color
  plot_limits = c(-1, 0.75), 
  axis_text_size = 4,
  is_category = TRUE) + 
  interior_annotation(label = "b3", cex = 2) +
  scale_color_manual(values = c("#8dd3c7", "#fdb462", "#bebada"))

### 2025-05-14 appendix.tex
Started updating the appendix

\[
\frac{2(r_2 + qr_2)}{\sqrt{3}(b_2 - 1)} = \frac{(1 + q)}{(b_1 - 1)}.
\] After rearranging this,

@@

\subsection{Single-cell gene
expression}\label{single-cell-gene-expression}

\subsubsection{Comparison with results of scDEED
recommendations}\label{comparison-with-results-of-scdeed-recommendations}

@@

\caption{\label{fig-pbmc-mse-umap}Assessing which of the two layouts
with UMAP and tSNE with different hyper-parameter setting (n\_neighbors:
\(30\), min\_dist: \(0.3\) (red); n\_neighbors: \(30\) (blue)) on the
PBMC data is the better representation using MSE for varying binwidth
(\(a_1\)). Colour used for the lines and points in the top plot and in
the scatterplots represents UMAP and tSNE layouts (a, b). Of the two,
layout b is optimal across all binwidths making it the best choice.}

>>>>

\[
\frac{2(r_2 + qr_2)}{\sqrt{3}(b_2 - 1)} = \frac{(1 + q)}{(b_1 - 1)}.
\]

After rearranging this,

@@

\section{Single-cell gene expression: comparison with results of scDEED
recommendations}\label{single-cell-gene-expression-comparison-with-results-of-scdeed-recommendations}

@@

\caption{\label{fig-pbmc-mse-umap}Assessing which of the two layouts
with UMAP and tSNE with different hyper-parameter setting (n\_neighbors:
\(30\), min\_dist: \(0.3\) (red); n\_neighbors: \(30\) (blue)) on the
PBMC data is the better representation using RMSE for varying binwidth
(\(a_1\)). Colour used for the lines and points in the top plot and in
the scatterplots represents UMAP and tSNE layouts (a, b). Of the two,
layout b is optimal across all binwidths making it the best choice.}
